Leadership across cultures and contexts: an exploratory analysis

### 2020-06-23 leadership_across_cultures_contexts.Rmd
Output chages
Edits to reduce

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010], many social scientists have endevored to obtain data from "non-WEIRD" populations. Dichotomizing populations, their psychologies, or social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially more diversity within "non-WEIRD" populations than within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that within a population all members are essentially similar, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community or political leaders, often overlooking leadership dynamics across contexts and levels of social organization, such as among kin and social groups. A second outstanding question is what benefits *and* costs both leaders and followers incur. While the benefits leaders accrue have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2020]. Leaders also play important roles in ritual contexts, including initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including conflict and cooperation, are also common among nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation-state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2020]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

@@

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with a high degree of social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideologies, and control of information [@cohen_comparative_1967; @bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are likely to be able to bear the costs of leadership more so than their counterparts [@glowacki_leadership_2015].

<!-- BENEFITS: DECISIONS IN THEIR INTEREST -->
Leadership positions are also commonly associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Many leadership positions in industrialized societies are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also generally viewed as sexually attractive, particularly to women [@buss_strategies_2006; @stanik_who_2010]. Most measures of reproductive success are positively associated with male social status, independent of subsistence type or status measure [@von_rueden_mens_2016].

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIT -->
Although, empirically, leadership is *associated* with benefits, evolutionary models require that leaders cause or generate benefits. Potential mechanisms include that leaders (1) facilitate collective actions that yield net benefits to themselves or close kin; (2) claim a fee or greater share of returns for their services; (3) receive reciprocal exchange in other currencies; or (4) gain other social or reproductive benefits by signaling their high qualities [@von_rueden_leadership_2015]. Leaders can also promote policies that either align with their individual interests or that are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019], even if 'just for a while.'

@@

Our first goal was to assess if discovery of evidence for each leadership dimension was biased by publication date, total pages of ethnography for a culture, and sex of the ethnographer(s). Our second goal was to assess the proportion of text records and cultures that provided evidence for each of coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, fit with the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

Our third goal was to assess the extent to which each leadership dimensions varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors entered as predictors, and author nested within culture as random intercepts. To investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions, we used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful when the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and k-nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the SI for descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of evolutionary theories of leadership. Guided by empirically derived features, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

The geographic distribution of the cultures sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` documents describing `r length(unique(all_data$d_culture))` cultures.

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by document publication date, female gender of any author or co-author, or total pages of ethnography from each culture. We found minimal evidence of bias, and therefore do not control for bias variables in subsequent analyses. For details, see the *Bias* section of the SI.

@@

For all `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (see Methods section). See Figure \@ref(fig:varsupport) and Table S\@ref(tab:glmmcoefs).

@@

The most common leader qualities were having high social status (over 90% of cultures), being experienced or accomplished, and being knowledgeable or intelligent (approximately 80% of cultures). Notably rare qualities included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

The random intercepts indicated that support for many variables did vary modestly by ethnographer and culture. Most of the standard deviations of the random author and culture intercepts for each model were between 0 and 1, on the scale of the linear predictor (Figures S\@ref(fig:FunctionRE) and S\@ref(fig:QualityRE)). In a few cases, one or the other was exactly 0, which can occur in models like ours that have many random effect levels (e.g., `r length(unique(all_data$author_ID))` authors) and a relatively small data set [@bates2015].

@@

We investigated four possibles sources of variation in evidence for our `r length(all_study_vars)` dimensions of leadership: group context, subsistence strategy, continental region, and leader sex. Specifically, we fit separate multiple logistic mixed effects regression models of each leadership dimension as a function of these four factors, with random intercepts for author nested within culture. We then compared the AIC value of each model to a model with only the fixed and random intercepts. Models whose AIC values were at least two less than the intercept-only models were deemed to outperform the intercept-only model [@burnham2002].

@@

df_compare2 %>% 
  filter(AIC_diff >= -2 & Estimate > 0.6) %>% 
  mutate(
    Type = map_chr(var, ~ str_split(., '_')[[1]][[1]]),
    Type = str_replace(Type, '\\.', ' '),
    Type = str_to_title(Type),
    `Percent of cultures` = 100*Estimate
  ) %>% 
  select(Variable, Type, `Percent of cultures`) %>% 
  arrange(-`Percent of cultures`) %>% 
  kable(digits = 1, caption = "Candidate universal dimensions of leadership. These dimensions were common in the ethnographic record (appearing in at least 60% of cultures), and did not appear to meaningfully vary by continental region, subsistence strategy,  group context, or leader sex.")

@@

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions (Figure \@ref(fig:heatmapdrop1)).

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three leadership dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

*Coercive authority* was documented in `r coerciveauthority`% of cultures. Unlike other more frequent leader qualities, which were generally positive or prosocial traits, *Coercive authority* was the most frequent negative or dominance-style leader quality. Additionally, it was one of the few dimensions with a complementary dimension, *No coercive authority*, which was documented in `r nocoerciveauthority`% of cultures (Figure \@ref(fig:varsupport)). *No coercive authority* demonstrated variability by subsistence strategy, with hunter-gatherers providing relatively higher levels of evidence. See Figure S\@ref(fig:heatmapdrop1).

@@

Here we report the cluster analysis of dimensions that used the correlation distance metric ($1 - cor$), and the Ward agglomeration algorithm (for results from other methods, see the SI). Figure \@ref(fig:cluster)A displays a dendrogram from the cluster analysis of the `r length(quality_vars)` leader quality dimensions and Figure \@ref(fig:cluster)B displays a dendrogram from the cluster analysis of the `r length(function_vars)` leader function dimensions. Each cluster analysis includes two estimates of significance for how strongly each cluster within the dendrogram is supported by the data. We rely on the AU (Approximately Unbiased) *p* values, which are computed by multiscale bootstrap resampling and represented as percentages (e.g., clusters with AU values $\gt95$ are strongly supported, and the top-level clusters are automatically highlighted by rectangles).

@@

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster and the smaller of which we identified as a *Prosociality* cluster. Within the management cluster, we identified two moderately supported subclusters: a *Strategize* and an *Organize* subcluster.

We identify these seven named subclusters as "features" of leadership dimensions in the ethnographic record. In naming these features, we gave extra weight to dimensions that were strongly supported empirically (Figure \@ref(fig:varsupport)), therefore not all dimensions within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership reviewed in the introduction, with the possible exception of *Competencies*, a single feature that included at least two important theoretical elements: dominance and prestige. Because features were aggregated dimensions, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

There were modest or no correlation between aggregageted features (see Figure S\@ref(fig:featurecor)). We fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, reported in the SI.

@@

Shamans and other leaders with supernatural qualities were surprisingly common. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of *Shamanism* as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, a form of knowledge or expertise, and *Experienced/accomplished*). See Figure S\@ref(fig:shamanism).

@@

Motivated by our feature analysis, and the apparent importance of shaman leaders, we tentatively advance a scheme, similar to @von_rueden_roots_2014[] [and inspired by @kaplan_embodied_2003; @kaplan_evolutionary_2009; @borgerhoff_mulder_intergenerational_2009], whereby leaders possess some combination of cognitive, social, material, and somatic capital that they deploy to provide benefits and impose costs (for further justification, see the Discussion section). Hence, we should be able to find examples of leaders with various combinations of these six elements.

@@

This exploratory study systematically evaluated evidence for `r length(all_study_vars)` dimensions of leadership in a stratified random sample of the ethnographic record of `r nrow(leader_cult)` largely nonindustrial cultures including hunter-gatherers, pastoralists, horticulturalists, and agriculturalists. Results revealed universal dimensions of leadership as well as important variation by group context, subsistence type, continental region, and leader sex.

@@

We found strong evidence that several leadership dimensions were universal across cultures and across contexts within cultures. Leaders are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures, and function to resolve conflict, organize cooperation, and provide counsel or direction in over 70% of cultures (Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (Figure \@ref(fig:varsupportcb)).

@@

Some common leadership dimensions were also relatively invariant across group context, subsistence strategy, continental region, and leader sex. These dimensions included qualities, such as being knowledgeable or intelligent, wealthy, strategically nepotistic, and generous; and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean that all leaders have all these traits. Instead, we mean that there is evidence for these traits across a solid majority of cultures that do not vary much by key measures of population or group variation (i.e., the four factors) (Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport)).

@@

Followers similarly benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types.

>>>>

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since Henrich et al.'s [-@henrich_weirdest_2010] seminal WEIRD people paper, many social scientists have sought data from "non-WEIRD" populations. Dichotomizing populations, their psychologies, or social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially more diversity within "non-WEIRD" populations than within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. Comparative research must assess cultural variation without assuming that all members within a population are essentially similar, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community or political leaders, often overlooking leadership across contexts and levels of social organization, such as among kin and social groups. A second outstanding question is what benefits *and* costs both leaders and followers incur. While the benefits leaders accrue have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, community leadership is typically informal and emerges in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Leaders also play important roles in ritual and religeous contexts [@singh_cultural_2017; @winkelman_crosscultural_2020], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Leadership in between-group conflict and cooperation is also common across nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation-state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

@@

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with privlaged social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideologies, and control of information [@bodley_cultural_2011].

@@

## The benefits and costs of leadership across cultures and contexts
<!-- COSTS: OPPURTUNITY COSTS -->
Evolutionary models generally assume that, on average, the fitness benefits of leadership outweigh the fitness costs. The costs, however, have received less attention. Fulfilling group responsibilities and mediating conflicts can be physically and socially costly, and some decisions will upset some followers [@wiessner_power_2010]. In the absence of institutionalized authority, all group members are equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including physical aggression, gossip, and ridicule [@boehm_purposive_2008; @hess_gossip_2017]. Physically formidable individuals with strong social networks are better equipped to manage these costs than others [@glowacki_leadership_2015].

<!-- BENEFITS: DECISIONS IN THEIR INTEREST -->
Leadership positions are also commonly associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Many leadership positions in industrialized societies are associated with increased financial compensation [@tsui_income_2017]. High social status, wealth, intelligence, and leadership status are also generally viewed as sexually attractive, particularly to women [@buss_strategies_2006; @stanik_who_2010]. Most measures of reproductive success are positively associated with male social status, independent of subsistence type or status measure [@von_rueden_mens_2016].

<!-- BENEFITS: 'MODEL' OF POTENTIAL BENEFIT -->
Although, empirically, leadership is *associated* with benefits, evolutionary models require that leaders cause or generate benefits. Potential mechanisms include that leaders (1) facilitate collective actions that yield net benefits to themselves or close kin; (2) claim a fee or greater share of returns for their services; (3) receive reciprocal exchange in other currencies; or (4) gain other social or reproductive benefits by signaling their high qualities [@von_rueden_leadership_2015]. Leaders can also promote policies that either align with their individual interests or that are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019].

@@

Our first goal was to assess if discovery of evidence for each leadership dimension was biased by publication date, total pages of ethnography for a culture, and sex of the ethnographer(s). Our second goal was to assess the proportion of text records and cultures that provided evidence for each coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, using the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

Our third goal was to assess variation in each leadership dimensions by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors as predictors, and author nested within culture as random intercepts. To investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions, we used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful when the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included hierarchical cluster analysis, logistic PCA, and minimal spanning trees and k-nearest neighbors (mst-kNN). See the SI for additional details. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of evolutionary theories of leadership. Guided by empirically derived features, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

The geographic distribution of the culture sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` documents describing `r length(unique(all_data$d_culture))` cultures.

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leadership dimensions by document publication date, female gender of any author or co-author, or total pages of ethnography from each culture. We found minimal evidence of bias, and therefore do not control for bias variables in subsequent analyses. For details, see the *Bias* section of the SI.

@@

For all `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and computed the proportion of cultures with at least one text record providing support for that dimension (see Methods section). See Figure \@ref(fig:varsupport) and Table S\@ref(tab:glmmcoefs).

@@

The most common leader qualities were having high social status (over 90% of cultures), being experienced or accomplished, and being knowledgeable or intelligent (approximately 80% of cultures). Notably rare qualities included physical formidability, having many social contacts, and being feared (approximately 35% of cultures) and charisma and fairness (approximately 20% of cultures).

The random intercepts indicated that support for many variables did vary modestly by author and culture. Most of the standard deviations of the random author and culture intercepts for each model were between 0 and 1, on the scale of the linear predictor (Figures S\@ref(fig:FunctionRE) and S\@ref(fig:QualityRE)). In a few cases, one or the other was exactly 0, which can occur in models like ours that have many random effect levels (e.g., `r length(unique(all_data$author_ID))` authors) and a relatively small data set [@bates2015].

@@

We investigated four possibles sources of variation in evidence for our `r length(all_study_vars)` leadership dimensions: group context, subsistence strategy, continental region, and leader sex. Specifically, we fit separate multiple logistic mixed effects regression models of each leadership dimension as a function of these four factors, with random intercepts for author nested within culture. We then compared the AIC value of each model to a model with only the fixed and random intercepts. Models whose AIC values were at least two less than the intercept-only models were deemed to outperform the intercept-only model [@burnham2002].

@@

df_compare2 %>% 
  filter(AIC_diff >= -2 & Estimate > 0.6) %>% 
  mutate(
    Type = map_chr(var, ~ str_split(., '_')[[1]][[1]]),
    Type = str_replace(Type, '\\.', ' '),
    Type = str_to_title(Type),
    `Percent of cultures` = 100*Estimate
  ) %>% 
  select(Variable, Type, `Percent of cultures`) %>% 
  arrange(-`Percent of cultures`) %>% 
  kable(digits = 1, caption = "Candidate universal leadership dimensions. These dimensions were common in the ethnographic record (appearing in at least 60% of cultures), and did not appear to meaningfully vary by continental region, subsistence strategy,  group context, or leader sex.")

@@

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying leadership dimensions, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions (Figure \@ref(fig:heatmapdrop1)).

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three leadership dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For group context, region, and leader sex see Figures S\@ref(fig:heatmapgroups), S\@ref(fig:heatmapregion), and S\@ref(fig:heatmapsex).

@@

*Coercive authority* was documented in `r coerciveauthority`% of cultures. Unlike other more frequent leader qualities, which were generally positive or prosocial traits, *Coercive authority* was the most frequent negative or dominance-style quality. Additionally, it was one of the few dimensions with a complementary dimension, *No coercive authority*, which was documented in `r nocoerciveauthority`% of cultures (Figure \@ref(fig:varsupport)). *No coercive authority* demonstrated variability by subsistence strategy, with hunter-gatherers providing relatively higher levels of evidence. See Figure S\@ref(fig:heatmapdrop1).

@@

Here we report the cluster analysis of dimensions that used the correlation distance metric ($1 - cor$), and the Ward agglomeration algorithm (for results from other methods, see the SI). Figure \@ref(fig:cluster)A displays a dendrogram from the cluster analysis of the `r length(quality_vars)` leader quality dimensions and Figure \@ref(fig:cluster)B a dendrogram from the cluster analysis of the `r length(function_vars)` leader function dimensions. Each cluster analysis includes two estimates of significance for how strongly each cluster within the dendrogram is supported by the data. We rely on the AU (Approximately Unbiased) *p* values, which are computed by multiscale bootstrap resampling and represented as percentages (e.g., clusters with AU values $\gt95$ are strongly supported, and the top-level clusters are automatically highlighted by rectangles).

@@

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management cluster and the smaller of which we identified as a *Prosociality* cluster. Within the management cluster, we identified two moderately supported subclusters: *Strategize* and *Organize* subclusters.

We identify these seven named subclusters as "features" of leadership dimensions in the ethnographic record. In naming these features, we gave extra weight to dimensions that were strongly supported empirically (Figure \@ref(fig:varsupport)), therefore not all dimensions within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership, with the possible exception of *Competencies*, a single feature that included at least two important theoretical elements: dominance and prestige. Because features were aggregated dimensions, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

There were modest or no correlation between features (see Figure S\@ref(fig:featurecor)). We fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, reported in the SI.

@@

Shamans and leaders with supernatural qualities were surprisingly common. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of *Shamanism* as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, a form of knowledge or expertise, and *Experienced/accomplished*). See Figure S\@ref(fig:shamanism).

@@

Motivated by our feature analysis, and the apparent importance of shaman leaders, we tentatively advance a scheme, similar to @von_rueden_roots_2014[] [and inspired by @kaplan_embodied_2003; @kaplan_evolutionary_2009; @borgerhoff_mulder_intergenerational_2009], whereby leaders possess some combination of cognitive, social, material, and somatic capital that they deploy to provide benefits and impose costs (for further justification, see the Discussion). Hence, we should be able to find examples of leaders with various combinations of these six elements.

@@

This exploratory study systematically evaluated ethnographic evidence for `r length(all_study_vars)` leadership dimensions from a sample of `r nrow(leader_cult)` largely nonindustrial cultures including hunter-gatherers, pastoralists, horticulturalists, and agriculturalists. Results revealed universal dimensions of leadership as well as important variation by group context, subsistence type, continental region, and leader sex.

@@

We found strong evidence that several leadership dimensions were universal across cultures and across contexts within cultures. Leaders are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures, and function to resolve conflicts, organize cooperation, and provide counsel or direction in over 70% of cultures (Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (Figure \@ref(fig:varsupportcb)).

@@

Some common leadership dimensions were also relatively invariant across group context, subsistence strategy, continental region, and leader sex. These dimensions included the qualities, knowledgeable or intelligent, wealthy, strategically nepotistic, and generous; and functions, resolving conflicts and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean all leaders have all these traits. Instead, we mean there is evidence for these traits across a solid majority of cultures that do not vary much by key measures of population or group variation (i.e., the four factors) (Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport)).

@@

Followers similarly benefited from increased material and social resources, but also often lost these resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types.

For both leaders and followers, individuals stood to lose what they were able to gain (Figure \@ref(fig:varsupportcb)). @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous Iroquois female leader, whose life exemplified the risks to both leaders and followers:
